\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-06-13 writing/manuscript.Rmd
updated manuscript and figures with new figures

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting. We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).

@@

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980]. While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, terrain ruggedness, prey availability), we decided to only use these three proxies to produce results that are easily comparable across species. Each species' model had the same structure:

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average). Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B11), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. B12).

@@

There was relatively strong agreement between models with and without temperature (Figs. B3, and B13), but including temperature always resulted in better fits. All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B). Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig. \@ref(fig:hrsfs)). Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation. Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B14).

@@

\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs. \@ref(fig:t-dist), B9, and B10). Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars). However, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs. \@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes also varied spatially due to heterogeneity in projected climate change (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse scenarios, but the direction of change at each location also varied across scenarios (most visible in cougars).

```{r, eval=FALSE, echo=FALSE}
# local predictions
readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Best', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

# BC predictions
readRDS('../data/cc-hgam-bc-projections-albers.rds') %>%
  group_by(lab) %>%
  filter(grepl('Best', scenario)) %>%
  summarise(mean = mean(d)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)
```

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 90% prediction interval within the range. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes in movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
knitr::include_graphics('../figures/local-distance-2100.png')
```

We do not predict any of the species to exhibit stronger selection for their current range, irrespective of climate change scenario. Changes in RSS by 2100 were between 0% and -3% under the best-case scenario, but they ranged from -8% to -14% under the worst-case scenario (Fig. \@ref(fig:t-hrsfs)). As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted increase in approximately 67% of the province, including their current range. The increase was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.

```{r, eval=FALSE}
# find species that decrease in RSS in less than half of BC
library('dplyr')
readRDS('data/cc-rss-bc-projections-albers.rds') %>%
  group_by(lab, scenario) %>%
  summarize(p_decrease = mean(lambda < 1)) %>%
  filter(p_decrease < 0.5) %>%
  mutate(p_increase = 1 - p_decrease)
```

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent. Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions-quantiles.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.26 to 117). The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
knitr::include_graphics('../figures/local-rss-2100.png')
```

@@

Our ability to respond changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].

Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021]. Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

@@

>>>>

```{r abstract, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change. Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century. We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities. In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation. We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting. We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public British Columbia Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).

@@

We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980]. While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, terrain ruggedness, prey availability), we decided to only use these three proxies to produce results that are easily comparable across species. Each species' model had the same structure:

@@

Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource. A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average). Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures. Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods). Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1. Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. B13).

@@

There was relatively strong agreement between models with and without temperature (Figs. B3, and B14), but including temperature always resulted in better fits. All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B). Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig. \@ref(fig:hrsfs)). Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation. Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B15).

@@

\noindent The predicted changes in distance traveled varied across species in both direction and magnitude, but worse climate-change SSPs always corresponded to greater absolute changes (Fig. \@ref(fig:t-dist)). Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs. B9 and B10). Overall, absolute changes by 2100 under the best-case SSP were small (approximately 0% to +4%), while under the worst-case SSP absolute changes ranged from ~2% (grizzly bears) to ~24% (cougars). Notably, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs. \@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes also varied spatially due to heterogeneity in projected climate change (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).

```{r, eval=FALSE, echo=FALSE}
library('dplyr')

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Best', scenario), year > 2095) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)
```

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use. Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025. Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line. The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four SSPs, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
knitr::include_graphics('../figures/local-distance-2100.png')
```

Median RSS was projected to decrease over time within the each species' observed range, but changes were stronger under worse SSPs (Fig. \@ref(fig:s-hrsfs)). Decreases were most pronounced in the bottom fifth percentile and most severe for elk, southern mountain caribou, and mountain goats. Of all the species, only cougars showed a clear increase in RSS for areas within their current range. The change in RSS between 2025 and 2100 varied spatially for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of their observed habitats, although cougars were predicted to increase their selection for patches with higher altitude.

```{r, eval=FALSE}
library('dplyr')

readRDS('../data/cc-hrsf-projections.rds') %>%
  group_by(lab) %>%
  filter(grepl('Best', scenario), year > 2095) %>%
  summarise(mean = mean(l_median)) %>%
  mutate(mean = (mean - 1) * 100,
         decreased = mean < 0) %>%
  arrange(mean)

readRDS('../data/cc-hrsf-projections.rds') %>%
  group_by(lab) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
    summarise(mean = mean(l_median)) %>%
  mutate(mean = (mean - 1) * 100,
         decreased = mean < 0) %>%
  arrange(mean)
```

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS. Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86)."}
if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
knitr::include_graphics('../figures/local-rss-2100.png')
```

@@

Our ability to respond changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in western Canada will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].

Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

attraction for disconnected patches of high elevation could lead to fragmentation or traps

@@

HRSFs do not account for migratory behavior due to reproduction or predation risk
